---
title: "Rubik’s Cube Competition Prevalence and Correlation with 3x3 Speed"
author: "Lauren Phung"
subtitle: "Are the fastest speedcubers the ones who have the most competitions near them?"
date: December 7, 2025
date-format: long
format:
  html:
    code-fold: true
    code-tools: true
execute:
  warning: false
  message: false
---

# Introduction

The Rubik's Cube is a popular 6-sided twisty puzzle first invented in the 1974 by Erno Rubik. Since the late 2000s, the cube's popularity has made a comeback. Many people enjoy the competitive aspect of the cube, aiming to solve it as fast as possible. The current 3x3 single world record is 3.05 seconds by Xuanyi Geng from China. The current 3x3 average world record is 3.90 seconds by Yiheng Wang from China.

Official Rubik's Cube competitions are governed by the World Cube Association (WCA), an entirely volunteer run non-profit organization that was founded in 2004. More than 273,000 people have competed in WCA sanctioned competitions. Competitions are held every week worldwide, but the geographic distribution of competition certainly isn't even. The United States has had over 2,000 cube competitions while other countries have had less than 10 or even none.     

My project aims to determine the relationship between the density of competitions in a given country and the speed of a country's competitors. I have 2 main research questions: How are World Cube Association competitions geographically distributed on a country level? Is the prevalence of WCA competitions correlated to the speed of the competitors in the region?


# Materials and methods

My research questions will be answered with the WCA Results Export. This is a massive dataset that is updated every week. It contains data about each competitor, the number of competitions attended, results and rankings, citizenship, gender, records, and world championship podiums. It also includes information about every competition: location, the number of competitors, what events were held, podium finishers.

There are many metrics to determine 3x3 speed. The WCA records the competitors' single and average when they compete. The 'single' is the fastest single attempt for the puzzle. The average is done taken after 5 solves, with the arithmetic mean taken after the slowest and fastest times are removed. Speedcubers acknowledge that good singles may be achieved with luck, while fast averages require more skill. 

3x3 Speed for each country will be determined with 3 variables:

1.  Mean of the top 10 3x3 averages
2.  Number of 3x3 Podiums at World Championships
3.  Number of 3x3 average world records

These 3 variables will be compared to the number of competitions in each country with a linear regression model. 


## Libraries

```{r, message=F, warning=F}
library(tidyverse)
library(leaflet)
library(knitr)
library(kableExtra)
library(htmlwidgets)
library(widgetframe)
library(sf)
library(maps)
library(sfhotspot)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(countrycode)
library(plotly)
library(broom)
library(knitr)
knitr::opts_chunk$set(widgetframe_widgets_dir = 'widgets' ) 
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download the required data
Two TSVs are downloaded from the WCA Results Export. Competitions includes data on every competition. Results contains all results from all competitions.  
```{r}
# loads the tsv from the WCA Results Export
competitions_tsv <- read.delim("data/WCA_export_Competitions.tsv", header = TRUE)

# contains all results from all competitions
results_tsv <- read.delim("data/WCA_export_Results.tsv", header = TRUE)
```


## Dependent Variable: Competition Count Per Country
Competitions count per country is calculated off of the competitions_tsv and using the function 'count'. 
```{r}
# cleaning competitions_tsv
competitions_tsv <- competitions_tsv %>%
  select(id, name, countryId, latitude, longitude, cancelled, year) %>%  # selects the columns I want to keep
  filter(cancelled == 0) %>% #removes cancelled competitions 
  mutate(longitude = as.numeric(longitude), # converts from chr to numeric
    latitude = as.numeric(latitude)) %>%
  mutate(latitude = latitude / 1e6, # converts from microdegrees to degrees
    longitude = longitude / 1e6) %>%
  mutate(countryId = countrycode(countryId, origin = "country.name", destination = "country.name")) %>%
  mutate(iso3c = countrycode(countryId, origin = "country.name", destination = "iso3c")
  ) # adds iso country code

# creates a dataframe that counts the number of comps per country
counts <- competitions_tsv %>%
  count(countryId, name = "count") %>%
  mutate(iso3c = countrycode(countryId, origin = "country.name", destination = "iso3c")) %>%
  rename(comp_count = count)
```

## Independent Variable 1: Mean of the top 10 3x3 averages for each country
Since the results_TSV contains every single average done from every single competition, I had to filter this dataset to only the averages from the top 10 fastest people per country. First, I removed all non-personal best results from the dataset. Then, I grouped by country and sliced the top 10 lowest results. The mean average was calculated using 'summarize'. 
```{r}
# shows only the competitor's fastest average
results_top10 <- results_tsv %>%
  select(average, eventId, personName, personId, regionalAverageRecord, personCountryId) %>%
  filter(eventId == 333, average != -1, average != 0) %>% # keeps valid 3x3 results
  group_by(personName) %>%
  slice_min(order_by = average, n = 1, with_ties = FALSE) %>%
  ungroup()

# grabs only the top 10 averages per country 
results_top10 <- results_top10 %>%
  group_by(personCountryId) %>%
  slice_min(order_by = average, n = 10) %>%
  ungroup()

#calculating mean average for each country
country_means <- results_top10 %>%
  group_by(personCountryId) %>%
  summarize(mean_average = mean(average)) %>%
  ungroup() %>%
  mutate(iso3c = countrycode(personCountryId, origin = "country.name", destination = "iso3c")) %>%
  mutate(mean_average = mean_average / 100)

```


## Independent Variable 2: Calculate number of 3x3 podiums at World Championships
Next, the number of 3x3 podiums at World Championships was calculated for each country. I filtered the results_tsv to only world championship competitions and further filtered this down to only look for top 3 placements in the final round as that is how the WCA defines a podium. 
```{r}
# filters for results at World Championship competitions
podiums <- results_tsv %>%
  select(pos, average, competitionId, eventId, roundTypeId, personName, personId, personCountryId) %>%
  filter(competitionId %in% c("WC2007", "WC2009", "WC2011", "WC2013", "WC2015", "WC2017", "WC2019", "WC2023", "WC2025")) %>% #filters only world championships
  filter(eventId == 333) %>% # keeps 3x3 only
  filter(roundTypeId == "f") %>% # final round only
  filter(pos <= 3) # top 3 placements are podiums

#counts number of podiums per country
podiums_per_country <- podiums %>%
  group_by(personCountryId) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  rename(podium_count = count) %>%
  mutate(iso3c = countrycode(personCountryId, origin = "country.name", destination = "iso3c"))

```

## Independent Variable 3: Calculating Number of 3x3 average world records
Next, the number of 3x3 average world records was calculated per country. World records were filtered from all results and then counted per country. 
```{r}
# filters for 3x3 world records
WRs <- results_tsv %>%
  select(eventId, personName, personId, regionalAverageRecord, personCountryId) %>%
  filter(eventId == 333) %>% # keeps 3x3 results
  filter(regionalAverageRecord == "WR") # world records only

#counts number of 3x3 average world records per country
WRs_per_country <- WRs %>%
  group_by(personCountryId) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  rename(WR_count = count) %>%
  mutate(iso3c = countrycode(personCountryId, origin = "country.name", destination = "iso3c"))# adds country code
  
```
## Combining all variables
All of the independent variables were joined together with the dependent variable into 1 dataframe. 
```{r}
full_df <- counts %>%
  left_join(country_means, by = "iso3c") %>%
  left_join(podiums_per_country,by = "iso3c") %>%
  left_join(WRs_per_country, by = "iso3c") %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% # adds 0 to countries that were missing, as missing countries have 0 podiums and 0 world records
  select(countryId, comp_count, mean_average, podium_count, WR_count) %>%
  mutate(countryId = replace_na(countryId, "Kosovo")) # the country code for Kosovo did not match 
  
```

# Results
## Competition Count by Country
```{r}
# loads world sf geometry - country borders
world <- ne_countries(scale = "medium", returnclass = "sf")

# corrects the missing iso country codes
world <- world %>%
  mutate(iso_a3 = case_when(
    iso_a3 == "-99" & name == "France" ~ "FRA",
    iso_a3 == "-99" & name == "Norway" ~ "NOR",
    iso_a3 == "-99" & name == "Somaliland" ~ "SOM",
    iso_a3 == "-99" & name == "Kosovo" ~ "XKX",
    TRUE ~ iso_a3)) %>%
  rename(iso3c = iso_a3)

world_data <- world %>%
  full_join(counts, by = "iso3c")

#creates color palette
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = world_data$comp_count,  
  na.color = "#e0e0e0"
)

# Create the choropleth map
leaflet(world_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(comp_count),
    fillOpacity = 0.7,
    color = "white",
    weight = 1,
    label = ~paste0(name, ": ", comp_count),  
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9)) %>%
  addLegend(
    pal = pal,
    values = ~comp_count,  
    opacity = 0.7,
    title = "Competition Count by Country",  
    position = "bottomright")
```
The interactive choropleth map below shows the dominance that the United States has over competition count.

```{r}
top_countries <- counts %>%
  select(comp_count, countryId) %>% 
  arrange(desc(comp_count)) %>%   # sort from highest to lowest
  head(10)

kable(top_countries, 
      col.names = c("Country", "Competition Count"),
      caption = "Top 10 Countries by Number of Competitions")

```
The United States has hosted 2,065 competitions. China, Brazil, India, and Australia round out the top 5. 

## Heat Map
```{r, message = F, warning = F, results = "hide"}
ggplot() +
  geom_sf(data = world, fill = NA, color = "black", linewidth = 0.3) +
  geom_density_2d_filled(data = competitions_tsv, aes(x = longitude, y = latitude), alpha = 0.8) +
  geom_density_2d_filled(alpha = 0.8) +
  geom_point(size = 0.5, alpha = 0.3, color = "white") +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal() +
  labs(
    title = "WCA Competitions Density Heatmap",
    x = "Longitude",
    y = "Latitude",
    fill = "Density"
  ) +
  coord_sf()
```
The heat map reveals that most competitions are clustered in the northeast region of the United States and Central Europe (Germany, Czechia, Austria). 

## Linear Regression - Mean of top 10 3x3 averages per country and competition count
Below are the results of the linear regression between mean of top 3x3 averages per each country and competition count. 
```{r}
#linear regression model
model1 <- lm(comp_count ~ mean_average, data = full_df)
summary(model1)

```
There is a negative relationship between, the mean of the top 10 3x3 averages and competition count. For every 1 second increase in top 10 3x3 average per country, competition count decreases by 6.22. The regression result is significant at the 0.05 level. However, the R-squared of the model is very low, which means that it does not do a good job of explaining the variation. 
```{r}
# scatter plot
plot1 <- ggplot(full_df, aes(x = mean_average, y = comp_count)) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_point(aes(text = paste("Country:", countryId, 
                              "<br>X:", round(mean_average, 2), 
                              "<br>Y:", comp_count)), 
             alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Competition Count vs 3x3 Speed (log scale)",
    x = "Mean 3x3 speed (seconds)",
    y = "Competition Count") +
  theme_minimal()

# adds interactive component
ggplotly(plot1, tooltip = "text")

```
This interactive scatter plot using log scale and a fitted regression line clearly shows the negative relationship between mean of the top 10 3x3 averages and competition count. The inverse of this relationship is true and easier to interpret: countries with faster 3x3 averages have more competitions. 

## Linear Regression - World Championship podiums per country and competition count
Below are the results of the linear regression between World Championship podiums per each country and competition count.
```{r}

top_podium_countries <- full_df %>%
  select(podium_count, countryId) %>% 
  arrange(desc(podium_count)) %>%   # sort from highest to lowest
  head(10)

kable(top_podium_countries, 
      col.names = c("Podium Count", "Country"),
      caption = "Top 10 Countries by World Championship Podium Count")
```
10 countries have achieved 3x3 podiums at World Championships. Poland and the United states are tied for first with 5 podiums each. 

```{r}
model2 <- lm(comp_count ~ podium_count, data = full_df)
summary(model2)

```
World championship podiums are positively correlated with competition count. With each additional world championship podium, the competition count increases by 149 competitions. This result is significant at below the 0.001 level. This model has better fit than the between the mean of top 10 3x3 averages and competition count. 

## Linear Regression - 3x3 average world records per country and competition count
Below are the results of the linear regression between 3x3 average world records per each country and competition count.
```{r}
top_wr_countries <- full_df %>%
  select(WR_count, countryId) %>% 
  arrange(desc(WR_count)) %>%   # sort from highest to lowest
  head(9)

kable(top_wr_countries, 
      col.names = c("World Record Count", "Country"),
      caption = "Top 9 Countries by World Record Count")
```
Only 9 countries have achieved a 3x3 average world record. Australia leads this category with 13 world records. China is a close second with 11 world records. 

```{r}
model3 <- lm(comp_count ~ WR_count, data = full_df)
summary(model3)

```
3x3 World records are positively correlated with competition count. With each additional world record, the competition count increases by 51 competitions. This result is significant at below the 0.001 level. 

# Conclusions

Overall, 3x3 speed is correlated with competition count. Countries with faster top 10 mean 3x3 averages have more competitions. Countries that have achieved more world championship podiums have more competitions. Countries that have achieved more world records have more competitions. 

The United States has by far the most competitions. The results of my linear regression models make a lot of sense as the World Cube Association was founded in the United States. Competitions are entirely volunteer run, so the United States was the first country to have organizers and WCA Delegates that hosted competitions. Organizers and Delegates need a lot of experience to run competitions, and the United States has that advantage over other countries. This also allowed the United States to develop a cubing culture early on, as cubing spreads locally through cube clubs and seeing other people solving cubes. 

This allowed the United States to grow cubing earlier and faster than other countries. With a larger amount of competitors, the fastest subset (10 people) of that larger amount is going to be faster at solving 3x3s than a country with a smaller number of competitors.

World championship podiums can only be achieved every 2 years. The world championship competition is hosted by a country that bids (similar to the Olympics). Traveling to a multi-date world championship is often time consuming and expensive, which makes it inaccessible to some competitors even if they are talented enough to achieve a podium. This is why the top countries that have achieved world championship podiums are all relatively developed countries. 

World records can be broken at any competition, but having more competitions nearby may motivate competitors to practice more. Having more competitions nearby also gives competitors a greater number of chances to break the world record. The top country for 3x3 world records is Australia because of the dominance of Feliks Zemdegs, who held the 3x3 world record and broke it 13 times from 2010 - 2019. 


# References

WCA Results Export - https://www.worldcubeassociation.org/export/results
